Precise spiking motifs in neurobiological and neuromorphic data

### 2022-10-26 manuscript.tex
Precise spike timing

A natural candidate to use these precise temporal patterns in the brain is to use Spiking Neural Networks (SNNs)~\citep{maass_networks_1997}. The approach which is at present most prominent in the SNNs community is to use existing algorithms from machine learning and to adapt them to the specificity of spiking architectures. One such example is to adapt the successes of deep learning algorithms and to transfer the back-propagation algorithm to SNNs, for instance with a surrogate gradient and try to ``cross-compile'' classical Neural Network to a spiking architecture. This approach is quite successful, and SNNs approach in some case the performance of Deep Learning algorithms, for instance on the N-MNIST dataset for categorizing digits in a stream of events. However, most biological neural systems use spikes and are obviously more efficient than current state-of-the-art vision systems, both in terms of efficiency (accuracy), in speed (latency), and energy consumption. There is therefore an immense gap in the way we understand biology to translate it to the efficiency of SNNs.  To go beyond the state-of-the-art, we will review here on one core computation of a spiking neuron, that is, is its ability to switch from the classical integrator mode (summing analog currents on its synapses) to a synchrony detector where it emits a spike whenever presynaptic spiking inputs are synchronized. To overcome the diversity of input presynaptic patterns, we will explore different existing architectures to learn to detect stable ``polychronous`` events, that is, volleys of spikes which are stable up to certain synaptic delays. These models will be compared in light of neuroscientific and computational perspectives. We review in this section theoretical and computational foundations of PG detection in models. In particular, our objective is to describe methods which fully exploit the capacity of spiking neurons to detect synchronous patterns.

@@

First, we will review models for the detection of polychronous groups.

@@

In particular, we will focus on the hypothesis that there exists at the scale of an assembly of neurons a representation based on a set of motifs defined by different relative spike times. %Here, we will review current literature on the detection of such motifs in generic raster plots. It is work in progress, where anybody interested can \emph{openly} join.

@@

We will then present some models for the detection of such motifs in arbitrary raster plots, biological or artificial (notably from event-based cameras). In particular, we will discuss models which exploit the variety of synaptic delays on the dendritic tree. Then, we will try to outline some possible strategies for learning these patterns and finally discuss possible perspectives.

@@

A notable exception is the polychronization model of Izhikevich~\citep{izhikevich_polychronization_2006}, which combined the construction of a random recurrent model of spiking neurons including such delays and whose weights evolved with a Spike-Time Dependent Plasticity (STDP) learning rule. In this model, raster plot analysis showed repeated activation of Polychronous Groups (PGs), i.e., specific spike patterns with a specific sequence of activations. In neuronal models, an efficient use or detection of these spatio-temporal patterns embedded in the spike train comes with the integration of heterogeneous delays~\citep{guise_bayesian_2014,zhang_supervised_2020}. Notably, Izhikevich~\citep{izhikevich_polychronization_2006} introduced the notion of polychronous groups (PGs) as a repetitive motif of spikes defined by a subset of neurons with different, yet precise, spiking delays. This representation has a much greater information capacity in comparison to other neural coding approaches through their connectivity and the possible coexistence of numerous superposed PGs.

>>>>

A natural candidate to use these precise temporal patterns in the brain is to use Spiking Neural Networks (SNNs)~\citep{maass_networks_1997}. The approach which is at present most prominent in the SNNs community is to use existing algorithms from machine learning and to adapt them to the specificity of spiking architectures. One such example is to adapt the successes of deep learning algorithms and to transfer the back-propagation algorithm to SNNs, for instance with a surrogate gradient, and try to ``cross-compile'' classical Neural Network to a spiking architecture. This approach is quite successful, and SNNs approach in some case the performance of Deep Learning algorithms, for instance on the N-MNIST dataset for categorizing digits in a stream of events. However, most biological neural systems use spikes and are obviously more efficient than current state-of-the-art vision systems, both in terms of efficiency (accuracy), in speed (latency), and energy consumption. There is therefore an immense gap in the way we understand biology to translate it to the efficiency of SNNs.  To go beyond the state-of-the-art, we will focus here on one core computation of a spiking neuron, that is, is its ability to switch from the classical integrator mode (summing analog currents on its synapses) to a synchrony detector where it emits a spike whenever presynaptic spiking inputs are synchronized. To overcome the diversity of input presynaptic patterns, we will explore different existing architectures to learn to detect stable ``polychronous`` events, that is, volleys of spikes which are stable up to certain synaptic delays. These models will be compared in light of neuroscientific and computational perspectives. We review in this section theoretical and computational foundations of PG detection in models. In particular, our objective is to describe methods which fully exploit the capacity of spiking neurons to detect synchronous patterns.

@@

% First, we will review models for the detection of polychronous groups.

@@

% In particular, we will focus on the hypothesis that there exists at the scale of an assembly of neurons a representation based on a set of motifs defined by different relative spike times. %Here, we will review current literature on the detection of such motifs in generic raster plots. It is work in progress, where anybody interested can \emph{openly} join.

@@

% We will then present some models for the detection of such motifs in arbitrary raster plots, biological or artificial (notably from event-based cameras). In particular, we will discuss models which exploit the variety of synaptic delays on the dendritic tree. Then, we will try to outline some possible strategies for learning these patterns and finally discuss possible perspectives.

@@

A notable example of a SNN using precise spike timings is the polychronization model of Izhikevich~\citep{izhikevich_polychronization_2006}, which combined the construction of a random recurrent model of spiking neurons including such delays and whose weights evolved with a Spike-Time Dependent Plasticity (STDP) learning rule. In this model, raster plot analysis showed repeated activation of Polychronous Groups (PGs), i.e., specific spike patterns with a specific sequence of activations. In neuronal models, an efficient use or detection of these spatio-temporal patterns embedded in the spike train comes with the integration of heterogeneous delays~\citep{guise_bayesian_2014,zhang_supervised_2020}. Notably, Izhikevich~\citep{izhikevich_polychronization_2006} introduced the notion of polychronous groups (PGs) as a repetitive motif of spikes defined by a subset of neurons with different, yet precise, spiking delays. This representation has a much greater information capacity in comparison to other neural coding approaches through their connectivity and the possible coexistence of numerous superposed PGs.
